---
title: "Workshop - Week 10"
---

This week we consider PCA and the spiked covariance model.

- PCA: classic and a high-dimensional example.
- Distribution of the largest eigenvalue of a sample covariance matrix
- The Tracy-Widom distribution of order 1
- Solve ODE to obtain Tracy-Widom F1 Distribution
- The spike covariance model

# Principal component analysis

First, let's review principal component analysis.

## PCA on a low-dimensional data set using the inbuilt function

We can start with the `iris` dataset that is built into R. The inbuilt function `princomp` performs a PCA on a dataset.

I am going to operate on this subset of the iris data:

```{r}
iris[-5]
```

```{r}
dim(iris[-5])
```

We perform the PCA on the dataset.

```{r}
irispca = princomp(iris[-5])
irispca
```


We can next show the model summary and see that the first component (Comp. 1) explains a large proportion of the variance. 

```{r}
summary(irispca)
```

```{r}
sqrt(0.9246187)
```

We can see the standard deviations of each of the principal components.
```{r}
irispca$sdev
```

Typically, a PCA de-means the data before performing the orthogonalisation algorithm. We can print the removed means with:
```{r}
irispca$center
```

We can extract the loadings (i.e., a matrix whose columns contain the eigenvectors).
```{r}
loadings(irispca)
```

We can plot a screeplot of the PCA. The default uses a boxplot.

```{r}
screeplot(irispca)
```

We can do something more similar to the lecture by specifying the type of plot and setting some graphics parameters.

```{r}
screeplot(irispca, type='lines', pch=19)
```

## PCA on a low-dimensional data set using our own function

We can implement our own PCA quite easily. First we setup the data:

```{r}
data = iris[-5]
```

```{r}
dim(data)
```

We compute the (vector) mean.
```{r}
mu = colMeans(data)
print(mu)
```

Notice this matches:
```{r}
irispca$center
```

We now subtract these means from the data.
```{r}
data = sweep(data, 2, mu)
```

```{r}
round(colMeans(data))
```

We now compute the covariance.
```{r}
S = cov(data)
```

We see that $p=4$ for the covariance matrix (as expected).
```{r}
dim(S)
```

We perform an eigendecomposition:
```{r}
E = eigen(S)
```

```{r}
E$values
```

We can compute the proportion of the total variance explained by the components using the eigenvalues.
```{r}
propvar = c()
total = sum(E$values)
for (lambda in E$values) {
  propvar = c(propvar, (lambda / total))
}
```

```{r}
propvar
```

Cumulative proportion of total variance:

```{r}
cumsum(propvar)
```

The "loadings" are simply given by
```{r}
round(E$vectors, 3)
```

Notice this matches:
```{r}
loadings(irispca, cutoff = 0.01)
```

```{r}
plot(E$values, type="b", pch=16)
```

## Higher-dimensional example

We are going to look at a higher-dimensional example: the Fashion MNIST dataset. We download the data.

```{r include=FALSE}
library(R.utils) # for gunzip
URL = "http://fashion-mnist.s3-website.eu-central-1.amazonaws.com/"
fn = "t10k-images-idx3-ubyte.gz"
download.file(paste0(URL, fn), fn)
gunzip(fn, overwrite=TRUE)
```

We need our code to read the data.
```{r}
read_idx = function(file_name, endian = "big")
{ if (!is.character(file_name)) stop("File_name must be character")
  file = file(file_name, "rb")
  # Check that the 2 first bytes contain the magic number  
  magic_1 = readBin(file, "raw", n = 1, size = 1, endian = endian)
  magic_2 = readBin(file, "raw", n = 1, size = 1, endian = endian)
  if (any(as.character(c(magic_1, magic_2)) != c("00", "00")))
    stop("The magic number doesn't match") 
  # Detect the type of data
  signed = TRUE
  type = as.character(readBin(file, "raw", n = 1, size = 1, endian = endian))
  if (type == "08")
  { size = 1
    signed = FALSE
    what = "integer"
  }
  else if (type == "09")
  { size = 1
    what = "integer"
  }
  else if (type == "0b")
  { size = 2
    what = "integer"
  }
  else if (type =="0c")
  { size = 4 
    what = "integer"
  }
   else if (type =="0d")
  { size = 4 
    what = "numeric"
  }
  else if (type == "0e")
  { size = 8
    what = "double"
  }
  else stop("Type byte found is not supported")
  # Get the number of dimensions
  number_dim = as.integer(readBin(file, "raw", n = 1, size = 1, endian = endian))
  # Get the actual dimensions
  dims = numeric(number_dim)
  for (i in seq_along(dims))
  { dims[i] = readBin(file, "integer", n = 1, size = 4, endian = endian)
  }
  
  cat(dims, "\n")
  
  # Read the data 
  n_records = prod(dims)
  data = readBin(file, what, n = n_records,  size = size, endian = endian, signed=signed)
  close(file)
  if (number_dim == 1) m = data
  else
  { m = array(data, dim=dims[number_dim:1])
    m = aperm(m)
  }
  return(m)
}
```

```{r}
rotate = function(x) t(apply(x, 2, rev))
```

```{r}
imshow = function(x, title = "", title.color = "black") {
  par(pty="s")
  image(rotate(x), axes = FALSE,
        col = hcl.colors(256),
        main = list(title, col = title.color),
        useRaster=TRUE)
}
```

```{r}
fn = "t10k-images-idx3-ubyte.gz"
X = read_idx(gsub(pattern = "\\.gz", "", fn))
```

```{r}
imshow(X[8,,])
```

```{r}
dim(X) = c(10000, 28*28)
```

```{r}
dim(X)
```

Generate the principal components.
```{r}
pca = princomp(X)
```

Do a screeplot of the first 20 eigenvalues.
```{r}
screeplot(pca, 20, type="lines", pch=16)
```

An alternative is to use the `prcomp` function which allows the "rank" parameter which allows us to specify a cut-off.

```{r}
pca = prcomp(X, rank=6)
```

```{r}
summary(pca)
```

The loadings are given by:
```{r}
dim(pca$rotation)
```

The mean is:
```{r}
mu = pca$center
dim(mu) = c(28, 28)
```

```{r}
imshow(mu)
```

We extract the first and second principal components and reshape them into images.
```{r}
pc1 = pca$rotation[,1]
dim(pc1) = c(28,28)

pc2 = pca$rotation[,2]
dim(pc2) = c(28,28)
```

We can also get the standard deviations of the principal components.

```{r}
pca$sdev[1]
```

```{r}
imshow(pc1)
```

```{r}
imshow(pc2)
```

# Distribution of the largest eigenvalue of a sample covariance matrix

We consider the distribution of the largest eigenvalue of a sample covariance matrix.

## Null case: $\Sigma = I_p$

We set:
```{r}
n = 100
p = 50
```

This gives:
```{r}
y = p/n
print(y)
```

And the upper support of the MP distribution is:
```{r}
b = (1+sqrt(y))^2
b
```

Eigenvalues given by `eigen` are in decreasing order, so we only need the first one. We write a function to extract this value.
```{r}
max.eigen = function(x) {
  eigen(x, only.values=TRUE)$values[1]
}
```

We define our (population) covariance matrix.
```{r}
Sigma = diag(1, p, p)
```

We can sample 10000 Wishart random matrices with degree of freedom $n$ and then extract the largest eigenvalue of each one.
```{r}
lambda1 = apply(rWishart(10000, n, Sigma), 3, max.eigen)
```

```{r}
length(lambda1)
```

We define the values $\mu_{np}$ and $\sigma_{np}$ (see lecture 10 p.3).

```{r}
mu.np = (sqrt(n-1/2)+sqrt(p-1/2))^2
sigma.np = (sqrt(n)+sqrt(p))*(1/sqrt(n-0.5)+1/sqrt(p-0.5))^(1/3)
```

$\mu_{np}$ is close to $b$ for large $p,n$.

```{r}
abs(mu.np - b)
```

```{r}
b
```

```{r}
c(mu.np, sigma.np)
```

```{r}
c(mean(lambda1), sd(lambda1))
```

We normalise the vector of $\lambda_1$'s using the formulas.

```{r}
F1 = (lambda1-mu.np)/sigma.np
```

We see that the mean is:
```{r}
mean(F1)
```

and the standard deviation:
```{r}
sd(F1)
```

We now plot the histogram.
```{r}
hist(F1, breaks=40, xlim=c(-4,4)) -> h
```

Replot the histogram with some nicer plot settings and we notice that it has a nice distribution.

```{r}
plot(h$mids, h$density, type="s", xlab="", ylab="", ylim=c(0,0.5))
title(main="Distribution of largest eigenvalue")
```

## The Tracy-Widom distribution of order 1

The Tracy-Widom distribution of order 1 is available in the package `RMTstat`. You can install with `install.packages('RMTstat')`

```{r include=FALSE}
library(RMTstat)
```

We use the library to plot the `dtw` against the histogram that we found.
```{r}
plot(h$mids, h$density, type="s", xlab="", ylab="", ylim=c(0,0.5))
curve(dtw, -5, 5, lwd=2, lty=3, col="darkmagenta", add=TRUE)
title(main="Distribution of largest eigenvalue vs. TW density")
```

The package `RMTstat` uses precomputed "magic" values to obtain the density.


## Solve ODE to obtain Tracy-Widom F1 Distribution

We solve the ODE on page 4 of Lecture 10.

For this next section, you'll need to install the `deSolve` and `gsl` packages. Do this with `install.packages(c('deSolve', 'gsl'))`.

Numerically solving to find the density of the Tracy-Widom distribution of order 1 is not easy; see the paper by Bornemann (2010).

We first follow Bejan (2005) and solve the ODE as a system of first-order ODEs using the package `deSolve` (a numerical ODE solver) and `gsl` for the special functions such as the Airy function.

```{r}
library(deSolve)
library(gsl)
```

```{r}
#See eq (22), (23) and (24) in 2014-Bejan.
deq = function(s, y, parms, ...) {
    list(c(y[2],
           s * y[1] + 2 * y[1] ^ 3,
           y[4],
           y[1] ^ 2,
           -y[1]))
}

# Setup time (backward)
t0 = 5
tn = -5
tspan = seq(t0, tn, length.out=1000)

# 'Initial' condition
# (actually a terminal condition since time goes backwards)

y0 = c(airy_Ai(t0),
        airy_Ai_deriv(t0),
        0,
        airy_Ai(t0) ^ 2,
        0)

# y0 = c(1.0834E-4, -2.4741E-4, 5.3178E-10, 1.1739E-8, 4.5743E-5)

# Solve the ODE

F.ode = ode(y=y0, times=tspan, func=deq, parms=NULL)
```

This allows us to plot:
```{r}
# time points
s = F.ode[, "time"]

# See eq (34) in 2014-Bejan
f1 = -0.5*(F.ode[, "4"]-F.ode[, "1"]) * exp(-0.5*(F.ode[, "3"]+F.ode[, "5"]))

plot(s, f1, type='l', main="Tracy-Widom F1 density", xlim=c(-5, 5), ylim=c(0,0.5), lwd=2, col="darkgreen")
lines(h$mids, h$density, type="s", xlab="", ylab="", ylim=c(0,0.5))
```

Alternatively, we can extract the solution for $q$ from the ODE solver and then use `approxfun` to create an interpolating function.
```{r}
q = approxfun(tspan, F.ode[, "1"])
plot(q, -5, 5, col = "darkgreen")
```

Then we use `q` and the definition for $I(s)$; see Eq (15) in 2005-Bejan.

```{r}
Ixx = function(s) {
  integrand = function(x) {
    (x - s)*q(x)^2
  }
  integrate(integrand, s, 5)$value
}
I = Vectorize(Ixx)
```

Then we use the definition for $I'(s)$; see below Eq (21).
```{r}
dIxx = function(s) {
  integrand = function(x) {
    -q(x)^2
  }
  integrate(integrand, s, 5)$value
}
dI = Vectorize(dIxx)
```

And then the definition for $J(s)$; see Eq (16).
```{r}
Jxx = function(s) {
  integrand = function(x) {
    q(x)
  }
  integrate(integrand, s, 5)$value
}
J = Vectorize(Jxx)
```

We plot them to check that they work.
```{r}
plot(I, -5, 5, col="darkgreen", ylim=c(-10,10))
curve(dI, -5, 5, col="darkred", add=TRUE)
curve(J, -5, 5, col="darkmagenta", add=TRUE)
```

Finally, we use the definition for $f_1(s)$ (see Eq (34)) and our functions above to generate the density and compare it to our histogram obtained from the simulation.

```{r}
f1 = function(s) {
  -0.5*(dI(s)-q(s)) * exp(-0.5*(I(s)+J(s)))
}

plot(f1, -5, 5, type='l', main="Tracy-Widom F1 density", xlim=c(-5, 5), ylim=c(0,0.5), lwd=2, col="darkmagenta")
lines(h$mids, h$density, type="s", xlab="", ylab="", ylim=c(0,0.5))
```

We can compare it to the density obtained by the RMTstat package.
```{r}
plot(f1, -5, 5, type='l', main="", xlim=c(-5, 5), lwd=2, col="green")
curve(dtw, -5, 5, lwd=2, lty=3, col="magenta", add=TRUE)
```

There are better techniques for numerically solving these Painlevé transcendents but they are harder to implement in R. See the paper by Bornemann (2010).


# The spike covariance model

We set:
```{r}
n = 100
p = 50
```

This gives:
```{r}
y = p/n
print(y)
```

Recall that a critical point is $1+\sqrt{y}$

```{r}
1+sqrt(y)
```

We define our spikes:
```{r}
alphas = c(5)
alphas
```

```{r}
alphas > 1+sqrt(y)
```

$\alpha_1$ should be a large fundamental spike that converges to (see p.15 Lecture 10):
```{r}
alphas[1] + y * alphas[1]/(alphas[1] - 1)
```

We define our spiked (population) covariance matrix.
```{r}
Sigma = diag(1, p, p)
for (k in 1:length(alphas)) {
  Sigma[k,k] = alphas[k]
}
```

We show our covariance matrix.
```{r}
imshow(Sigma)
```
```{r}
Sigma[1:5,1:5]
```
We are going to compare against the Marcenko-Pastur density.
```{r}
dmp = function(x, y, sigma=1) {
  a = (1-sqrt(y))^2
  b = (1+sqrt(y))^2
  ifelse(x <= a | x >= b, 0, suppressWarnings(sqrt((x - a) * (b - x))/(2 * pi * sigma * x * y)))
}
```

```{r}
eigenvalues = function(x) {
  eigen(x, only.values=TRUE)$values
}
```

We generate many eigenvalues
```{r}
lambdas = apply(rWishart(1000, n-1, Sigma)/(n-1), 3, eigenvalues)
```

```{r}
c(1000, p)
```

```{r}
length(lambdas)
```

We see there is an accumulation of eigenvalues around the fundamental spike and then the bulk is described by the MP distribution.
```{r}
hist(lambdas, breaks=1000, freq = FALSE)
curve(dmp(x, y=p/n), from = 0, to = 1.5*max(lambdas), lty=1, lw=2, col=2, add=TRUE)
abline(v = alphas[1] + y * alphas[1]/(alphas[1] - 1), lty=2, lwd=2, col='darkgreen')
```
