---
title: "Workshop - Week 2"
output:
  html_document:
    toc: yes
---

# Datasets

## Setup

Some of the data sets we will look at are stored in `idx` format, so we need to install some R libraries to read this format.

```{r}
# install.packages(c('R.utils'))
```

```{r}
library(R.utils) # for gunzip
```

We now load these libraries.

```{r}
read_idx = function(file_name, endian = "big")
{ if (!is.character(file_name)) stop("File_name must be character")
  file = file(file_name, "rb")
  # Check that the 2 first bytes contain the magic number  
  magic_1 = readBin(file, "raw", n = 1, size = 1, endian = endian)
  magic_2 = readBin(file, "raw", n = 1, size = 1, endian = endian)
  if (any(as.character(c(magic_1, magic_2)) != c("00", "00")))
    stop("The magic number doesn't match") 
  # Detect the type of data
  signed = TRUE
  type = as.character(readBin(file, "raw", n = 1, size = 1, endian = endian))
  if (type == "08")
  { size = 1
    signed = FALSE
    what = "integer"
  }
  else if (type == "09")
  { size = 1
    what = "integer"
  }
  else if (type == "0b")
  { size = 2
    what = "integer"
  }
  else if (type =="0c")
  { size = 4 
    what = "integer"
  }
   else if (type =="0d")
  { size = 4 
    what = "numeric"
  }
  else if (type == "0e")
  { size = 8
    what = "double"
  }
  else stop("Type byte found is not supported")
  # Get the number of dimensions
  number_dim = as.integer(readBin(file, "raw", n = 1, size = 1, endian = endian))
  # Get the actual dimensions
  dims = numeric(number_dim)
  for (i in seq_along(dims))
  { dims[i] = readBin(file, "integer", n = 1, size = 4, endian = endian)
  }
  
  cat(dims, "\n")
  
  # Read the data 
  n_records = prod(dims)
  data = readBin(file, what, n = n_records,  size = size, endian = endian, signed=signed)
  close(file)
  if (number_dim == 1) m = data
  else
  { m = array(data, dim=dims[number_dim:1])
    m = aperm(m)
  }
  return(m)
}
```

```{r}
source('read_idx.r')
```

## MNIST

This is a famous data set of handwritten digits. We will get the small version. We download the data into a local file.

```{r}
URL = "http://yann.lecun.com/exdb/mnist/"
fn = "t10k-images-idx3-ubyte.gz"
download.file(paste0(URL, fn), fn)
gunzip(fn, overwrite=TRUE)
```
We now read it into the `train` object.
```{r}
train = read_idx(gsub(pattern = "\\.gz", "", fn))
```

This dataset has the following dimensions.
```{r}
dim(train[1,,])
```

We create some functions to view these observations.
```{r}
rotate = function(x) t(apply(x, 2, rev))

imshow = function(x, title = "", title.color = "black") {
  image(rotate(x), axes = FALSE,
        col = grey(seq(0, 1, length = 256)),
        main = list(title, col = title.color),
        useRaster=TRUE)
}
```

```{r fig.height=1, fig.width=1}
imshow(train[5,,])
```

```{r}
par(mfrow=c(4, 4), mar=c(0, 0.2, 1, 0.2))

for (i in 1:16) {
  nr <- i * 12
  imshow(train[nr,,])
}
```

```{r}
28*28
```

```{r}
x = train[1,,]
```

```{r}
dim(x) = c(28*28)
```

```{r}
dim(train) = c(10000, 28*28)
```

```{r}
S = cov(train)
```

```{r}
mu = colMeans(train)
```

```{r}
dim(mu) = c(28, 28)
```

```{r}
imshow(mu)
```
```{r}
URL = "http://yann.lecun.com/exdb/mnist/"
fn = "t10k-labels-idx1-ubyte.gz"
download.file(paste0(URL, fn), fn)
gunzip(fn, overwrite=TRUE)
```
We now read it into the `train` object.
```{r}
label = read_idx(gsub(pattern = "\\.gz", "", fn))
```

```{r}
dim(x) = c(28, 28)
```

```{r}
imshow(x)
```

```{r}
label[1]
```

```{r}
mu = colMeans(train[label==7,])
```

```{r}
dim(mu) = c(28, 28)
```

```{r}
imshow(mu)
```
```{r}
S = cov(train[label==7,])
```

```{r}
imshow(S)
```


## Fashion MNIST

```{r}
URL = "http://fashion-mnist.s3-website.eu-central-1.amazonaws.com/"
fn = "t10k-images-idx3-ubyte.gz"
download.file(paste0(URL, fn), fn)
gunzip(fn, overwrite=TRUE)
```

```{r}
train = read_idx(gsub(pattern = "\\.gz", "", fn))
```

```{r}
dim(train)
```

```{r}
par(mfrow=c(4, 4), mar=c(0, 0.2, 1, 0.2))
for (i in 1:16) {
  nr <- i * 10
  imshow(train[nr,,])
}
```

```{r}
URL = "http://fashion-mnist.s3-website.eu-central-1.amazonaws.com/"
fn = "t10k-labels-idx1-ubyte.gz"
download.file(paste0(URL, fn), fn)
gunzip(fn, overwrite=TRUE)
```
We now read it into the `train` object.
```{r}
label = read_idx(gsub(pattern = "\\.gz", "", fn))
```


```{r}
dim(train) = c(10000, 28*28)
```

```{r}
mu = colMeans(train[label==8,])
```

```{r}
dim(mu) = c(28, 28)
```

```{r}
imshow(mu)
```


# Salient features of Big Data

## Heterogeneity

```{r}
set.seed(123)
mus <- rnorm(5, mean=0, sd=10)
sds <- rchisq(5, 1)
```

```{r}
mus
```

```{r}
library(MASS)
```

```{r}
Sigma = diag(sds)
print(Sigma)
```

```{r}
Z = mvrnorm(n=3000, mus, Sigma)
```

```{r}
dim(Z)
```

```{r}
X = as.vector(mvrnorm(n=3000, mus, Sigma))
hist(X, breaks=100)
```

```{r}
summary(X)
```

```{r}
mu <- mean(X)
sd <- sd(X)
hist(X, breaks=100, freq=FALSE, main="Bad model that doesn't capture heterogeneity")
curve(dnorm(x, mean=mu, sd=sd), col=2, lty=2, lwd=2, add=TRUE)
```

## Noise accumulation

Try to replicate the example (Fig. 1) in the "Noise Accumulation" section of the paper *Challenges of Big Data analysis* by Fan, Han, and Liu. Useful functions are `prcomp` and `mvrnorm` (in the `MASS` package).

```{r}
library(MASS)
```

We setup the number of observations `n` and the number of dimensions `p`.
```{r}
n <- 100
p <- 1000
```

The first class has a zero mean.
```{r}
mu <- rep(0, p)
```

The second class has mean `eta`. It is a sparse mean with 0 entries everywhere except for the first 10 dimensions where the mean is 3.
```{r}
eta = rep(0, p)
eta[1:10] = 3
```

```{r}
eta
```

Sample from each class.
```{r}
I <- diag(2,p,p)
X <- mvrnorm(n=n, mu, I)
Y <- mvrnorm(n=n, eta, I)
```

Plot the first two dimensions against each other for the two classes.
```{r}
m = 1
plot(X[,m], X[,m+1], bg=1, pch=19, xlim=c(-10,10), ylim=c(-10,10))
points(Y[,m], Y[,m+1], col="red", bg="red", pch=24)
```

Stack the data into one matrix.
```{r}
Z <- rbind(X,Y)
```

```{r}
dim(Z)
```

```{r}
m <- 1000
pcaZ <- prcomp(Z[,1:m])
pX <- pcaZ$x[1:n,]
pY <- pcaZ$x[(n+1):nrow(pcaZ$x),]
plot(pX[,1], pX[,2], bg=1, pch=19, xlim=c(-10,10), ylim=c(-10,10))
points(pY[,1], pY[,2], col=2, bg=2, pch=24)
```