---
title: "Workshop - Week 1"
---

\newcommand{\X}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\C}{\mathbf{C}}

```{r setup, include=FALSE}
par(family="sans", cex=0.8, cex.axis=1.0, cex.lab=1.0, cex.main=1.1, cex.sub=1.0)
```

# Crash course

This introduction workshop is inspired by code from Everitt's book *An Introduction to Applied Multivariate Analysis with R*. This is potentially a good read if you like the applied side of multivariate analysis.

## Multidimensional data basics

### Constructing vectors

Create a vector, using the command `c()` (for concatenate)

```{r}
x = c(1, 2, 3, 4)
x
```

```{r}
c(1, 2, 3, 4) -> x
x
```

Sum all elements of x:
```{r}
sum(x)
```

Square all elements of x:
```{r}
x^3
```

Get the third element of x:
```{r}
x[3]
```

```{r}
result = c()
result
```


Add an extra element:
```{r}
result = c()
for (i in 1:10) {
  result = c(result, i)
}
result
```

Create a vector using the command `seq()` (for sequence)
```{r}
x = seq(from=1, to=10, length.out=10)
x
```

Obtain the length of x:
```{r}
length(x)
```

```{r}
1:10
```

Create a vector using the command 'rep()' (for repeat)
```{r}
x = rep(0, 100)
x
```

```{r}
y <- rep(1:4, each = 3)
y
```

### Constructing matrices

Library for the matrix commands.

```{r}
library(Matrix)
```

Construct a matrix.

```{r}
A = matrix(c(1, 2, 3, 4, 5, 6), ncol=2, byrow=TRUE)
A
```

Access elements.

```{r}
A[,2]
```

Access rows.

```{r}
A[1,]
```

Access columns.

```{r}
A[,1]
```

Construct by column first.

```{r}
B = matrix(1:9, byrow=F, ncol=3)
print(B)
```

The `diag` function has multiple uses.
Construct a diagonal matrix from a vector.

```{r}
D = diag(rep(2,4))
D
```

Construct an identity matrix of a certain size.

```{r}
I = diag(4)
I
```

Extract or change the diagonal of a matrix.

```{r}
print(A)
```

```{r}
cat("1\t2\n")
cat("2\t3\n")
```

```{r}
B
```


```{r}
diag(B) = 1
```

```{r}
B
```

Construct a matrix of all ones, two possible approaches:

```{r}
ONES = matrix(rep(1, 9), ncol=3)
ONES
```

```{r}
ONES = matrix(1, nrow=3, ncol=3)
ONES
```

```{r}
ONES
```


### Basic operations

Create some vectors and matrices.

```{r}
x = c(1, 2, 3)
y = c(4, 5, 6)
z = seq(1, 10, by=1)
B = matrix(1:6, 2, 3)
```

```{r}
y
```


```{r}
class(y)
```

To make sure that R respects dimensions, turn them into matrices.

```{r}
x = as.matrix(x)
y = as.matrix(y)
```

```{r}
class(x)
```

```{r}
dim(x)
```

```{r}
x
```


#### Basic operations

Transpose operations.

```{r}
t(A)
t(B)
t(D)
```

Element-wise operations on matrices.

```{r}
A
```

```{r}
B
```

```{r}
A %*% B
```

```{r}
A+t(B)
A-t(B)
A*t(B) # CAUTION!!! This is not matrix multiplication.
A/t(B)
A^t(B)
```

Element-wise operations on vectors.

```{r}
x+y
x-y
x*y
x/y
y^x
```

#### Matrix and vector operations

This would give an error message: non-conformable.

```{r}
dim(A)
```

```{r}
dim(B)
```

```{r}
A %*% B
```

Check the matrix dimension.

```{r}
dim(A)
```

```{r}
dim(t(B))
```

A correct calculation is obtained by transposing the matrix $B$ so that it has the correct dimensions to be multiplied with $A$.

```{r}
A %*% B -> C
print(C)
```

Here are some alternatives that also work:

```{r}
A %*% B

B %*% A

B %*% A

x %*% t(y)

t(x) %*% y

# x %*% A
```

Multiplies each column of B by a number by using the diagonal matrix $D$.

```{r}
D = diag(c(1,2,3))
D
```

```{r}
B
```


```{r}
B %*% D
```

Multiplies each row of B by a number by using a diagonal matrix.

```{r}
diag(c(3,4)) %*% B
```

### Other operations

Determinant of a matrix.

```{r}
D
```

```{r}
det(D)
```

```{r}
det(ONES)
```

Inverse of a matrix $A x = b$ $x = A^{-1} b$

```{r}
Di = solve(D)
Di
```

$ A A^{-1} = I$

```{r}
D %*% Di
```

```{r}
Di %*% D
```

You can create an almost-singular matrix by choosing a small variance noise matrix `N` and see what happens with the inverse of `(1+N)`. This is to show that very small numerical errors can make it very hard to invert a matrix.

```{r}
set.seed(112312)
N<-matrix(rnorm(9, sd=10^-1),3,3)
print(N)
```

```{r}
ONES
```

```{r}
Ii = solve(ONES+N)
Ii
```

```{r}
(ONES+N) %*% Ii
```

```{r}
Ii%*%(I+N)
```

## Plotting multidimensional data

Visualisation is very important when exploring a new data set. Here are some useful ways to look at multi-dimensional data.

### Box plot

```{r}
mtcars
```

Standard boxplot.

```{r}
boxplot(mtcars[,1:7], main="Boxplot of car data")
```

### Scatter plot

Bivariate scatter plot.
```{r}
plot(mtcars$hp, mtcars$mpg)
```

Add labels along the axes
```{r}
plot(mtcars$hp, mtcars$mpg, xlab="Horse power", 
     ylab="Miles per gallon", main="Cars data")
```

All possible bivariate scatter plots.
```{r}
pairs(mtcars[,1:7], panel=panel.smooth,
      main="Scatterplot matrix of car data")
```

We can make a nicer version but we need to create two helper functions first. This one puts histograms on the diagonal.
```{r}
panel.hist <- function(x, ...) {
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(usr[1:2], 0, 1.5) )
  h <- hist(x, plot = FALSE)
  breaks <- h$breaks; nB <- length(breaks)
  y <- h$counts; y <- y/max(y)
  rect(breaks[-nB], 0, breaks[-1], y, 
       col = "cyan", ...)
}
```

This one puts (absolute) correlations on the upper panels, with size proportional to the correlations.
```{r}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...) {
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y))
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * r)
}
```

Now let's do it.
```{r}
pairs(mtcars[,1:7], panel=panel.smooth, 
      upper.panel=panel.cor, diag.panel=panel.hist)
```

### Bubble plots

```{r}
plot(mtcars$hp, mtcars$mpg, pch = 20, 
     xlab = "Horse power", ylab = "Miles per Gallon", 
     ylim=c(0,40))
```
We can add a third variable by using circles.
```{r}
plot(mtcars$hp, mtcars$mpg, pch = 20, 
     xlab = "Horse power", ylab = "Miles per Gallon", 
     ylim=c(0,40))
symbols(mtcars$hp, mtcars$mpg, circles = mtcars$cyl, 
        add = TRUE)
title("Bubble plot: The radius of the circle 
      indicates weight")
```

We can add even more variables with rectangles and colors:
```{r}
plot(mtcars$hp, mtcars$mpg, pch = 20, 
     xlab = "Horse power", ylab = "Miles per Gallon", 
     ylim=c(0,40))
symbols(mtcars$hp, mtcars$mpg, 
        rectangles = cbind(mtcars$wt,mtcars$drat), 
        fg=mtcars$cyl, lwd=2, add = TRUE)
title("Width: weight, height: rear axel ratio")
legend("topright", c("4 cyl", "6 cyl", "8 cyl"), 
       lty=1, col=c(4,6,8), lwd=2)
```

### Star plots

```{r}
stars(mtcars[, 1:7], key.loc = c(14, 2), scale=T,
      main = "Motor Trend Cars")
```

```{r}
stars(mtcars[, 1:7], key.loc = c(14, 2), scale=T,
      main = "Motor Trend Cars", full = FALSE)
```

Better approach, segment plot with colors.
```{r}
palette(rainbow(12, s = 0.6, v = 0.75))
stars(mtcars[, 1:7], key.loc = c(14,2), scale=T, 
      main = "Motor Trend Cars", 
      draw.segment=TRUE)
# with more control over position of labels:
loc <- stars(mtcars[, 1:7], key.loc = c(14,2), scale=T, 
       main = "Motor Trend Cars", 
       draw.segment=TRUE)

loc <- stars(mtcars[, 1:7], key.loc = c(14,2), scale=T, 
      labels=NULL, main = "Motor Trend Cars", 
      draw.segment=TRUE)
# loc contains the centers of the segment plots
# write name of car in the middle of each segment plot
text(loc[,1],loc[,2]-.8, 
     row.names(mtcars), 
     col = "black", cex = 0.6, xpd = TRUE)
palette("default") # set colors back to default
```

### Parallel coordinates plot

This is one of my favourites for high-dimensional data.

```{r}
library(MASS)
```

```{r}
parcoord(mtcars[,1:7])
```

Color can help.

```{r}
parcoord(mtcars[,1:7], col=mtcars$cyl)
```

## Random vectors and matrices

### Random matrix

```{r}
rnorm(6)
```


```{r}
# set.seed(12312)
x = matrix(rnorm(6), ncol=2)
x
```

Notice that `mean(x)` DOES NOT produce what we want.
```{r}
mean(x)
```

```{r}
dim(x)
```

```{r}
n = dim(x)[1]
n
```

```{r}
n = nrow(x)
n
```

```{r}
p = ncol(x)
p
```

Empirical mean.
```{r}
n = nrow(x)
ones = matrix(rep(1,n), ncol=1)
mu = t(x) %*% ones / n
print(mu)
```

```{r}
x
```

We can also use the apply function here.
```{r}
apply(x, 2, mean)
```

```{r}
apply(x, 1, mean)
```

```{r}
colMeans(x)
```

```{r}
rowMeans(x)
```


### Variance and standard deviation of a vector

```{r}
x
var(x[,1])
var(x[,2])

sd(x[,1])
sd(x[,2])

# covariance 
var(x[,1], x[,2])
```
```{r}
x
```


Variance-covariance matrix.

```{r}
var(x)
```

Correlation matrix.

```{r}
cor(x)
```

### Sample variance-covariance

3x3 matrix of 1s.

```{r}
ones
```


```{r}
ONES <- ones %*% t(ones)
ONES
```

Identity matrix.

```{r}
diag(3)
```

Matrix computation of S (unbiased)

```{r}
(1/(n-1)) * t(x) %*% (diag(3)-(1/n)*ones %*% t(ones)) %*% x
```

Produces the same result

```{r}
cov(x)
```

# Salient features of Big Data

## Outliers in higher dimensions are not obvious

### Outliers in univariate case

```{r}
set.seed(123)
dat <- matrix(rnorm(5*100),100,5)
summary(dat)
dat[23,4] <- dat[23,4] * 10
summary(dat)
pairs(dat)
```

Extract its location
```{r}
which.min(dat[,4])
```

### Multivariate outliers

```{r}
load(file = "simpleExample.rda")
```

There are no clear univariate outliers here
```{r}
par(mfrow=c(1,4))
plot(dat, xlim = c(-5,5), ylim = c(-5,5))
barplot(dat[,1], main="x values")
barplot(dat[,2], main="y values")
```
Using Mahalanobis distance
```{r}
d <- mahalanobis(dat, colMeans(dat), cov(dat))
barplot(d, main="Mahalanobis")
which.max(d)
```

Create chi-squared QQ-plot.
```{r}
par(mfrow=c(1,1))
qqplot(qchisq(ppoints(301), df = 2), d,
       main = expression("Q-Q plot for" ~~ {chi^2}[nu == 2]))
```

Now the outlier is clearly visible.
```{r}
par(mfrow = c(1,1))
plot(dat, xlim=c(-5,5), ylim=c(-5,5))
points(dat[301,1],dat[301,2],col="red")
```

# Distances between vectors in high dimension

```{r}
p = 50
n = 100
X = matrix(runif(n*p), nrow=n, ncol=p)
d = c()
for (i in 1:n) {
  for (j in i:n) {
    d = c(d, norm(X[i,]-X[j,], "2"))
  }
}
hist(d, 50)
```
