---
title: "Workshop - Week 11"
---

```{=tex}
\newcommand{\X}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\S}{\mathbf{S}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\C}{\mathbf{C}}
\newcommand{\1}{\mathbb{1}}
```
```{r}
rm(list = ls())
```

# The Markowitz portfolio

We consider the construction of financial portfolios of stocks using the Markowitz approach.

## Closed-form solution

Define the (population) returns of the stocks.

```{r}
mu = c(4.27, 0.15, 2.85)
names(mu) = c("S1", "S2", "S3")
```

Define the (population) covariance matrix.

```{r}
Sigma = matrix(c(1, 0.18, 0.11, 
                  0.18, 1.1, 0.26, 
                  0.11, 0.26, 1.99), nrow=3, ncol=3)

rownames(Sigma) = c("S1", "S2", "S3")
colnames(Sigma) = c("S1", "S2", "S3")
```

Show the correlations.

```{r}
cov2cor(Sigma)
```

Define the Markowitz function that solves the optimal portfolio problem given $(\mu, \Sigma)$.

```{r}
markowitz = function(mu, Sigma){
  
  InvSigma = try(solve(Sigma), TRUE)
  if("try-error" %in% class(InvSigma)) stop
  
  One = rep.int(1,length(mu))

  a = as.double(t(mu) %*% InvSigma %*% mu)
  b = as.double(t(mu) %*% InvSigma %*% One)
  c = as.double(t(One)%*% InvSigma %*% One)
  d = a*c - b^2
  
  Phi = (a/d)*(InvSigma %*% One) - (b/d)*(InvSigma %*% mu)
  Theta = (c/d)*(InvSigma %*% mu) - (b/d)*(InvSigma %*% One)
  
  R = b/c
  Pi = Phi + Theta*R
  
  sigma = Vectorize(function(r)  sqrt((c/d)*((r-(b/c))^2) + (1/c)))
  
  return(list(Pi=Pi, R=R, sigma=sigma))
}
```

We obtain the result using our function. This function returns the returns, the portfolio allocation, and a function that computes the risk.

```{r}
mp = markowitz(mu, Sigma)
```

We can extract the optimal return.

```{r}
mp$R
```

We can extract the optimal portfolio weights.

```{r}
mp$Pi
```

## Comparing to a convex optimisation problem solver

Our function is similar to solving the quadratic programming problems of the form $$\min \frac{1}{2} x' \Sigma\, x -\mathbb{1}' x$$ with the constraints $\mathbb{1}' x = 1$.

The function `solve.QP` is available with the `quadprog` package. Install it using `install.packages("quadprog")` and make it available in your R session.

```{r}
#install.packages("quadprog")
library("quadprog")
```

The function `solve.QP` solves problems of the form $$ \min \frac{1}{2} x' D x -d' x$$ with the constraints $A' x \ge b_0$. Here we set $D = \Sigma$, $d=\mathbf{1}$, $A = \mathbf{1}$, $d=0$, and $b_0 = 1$. This means we need to map our problem to this layout.

```{r}
d = matrix(0, ncol(Sigma), 1)
A = matrix(1, ncol(Sigma), 1)

Pi = solve.QP(Sigma, d, A, bvec=1, meq=1)$solution
```

We get the same optimal portfolio as our own solver (above).

```{r}
Pi
```

The portfolio return (in percent) is given by

```{r}
R = t(Pi) %*% mu
R
```

## Markowitz frontier

Our closed-form solver above also provided a `sigma` function that maps from return $R$ (in percent) to level of risk $\sigma_0$. We can use this to draw the Markowitz frontier.

```{r}
mp = markowitz(mu, Sigma)
returns = seq(0, 6, 0.1)
sigmas = mp$sigma(returns)
sigmas
```

Notice how we flip the axes and draw the function the other way around so it is one-to-one.

```{r}
plot(sigmas, returns, xlab=expression(sigma), ylab=expression(R), type = "l", main ="Markowitz frontier", xlim=c(0.7,1), col='darkmagenta', lwd=2)
```

## Real-world example

We can do a real-world example but we need to install a few packages.

```         
install.packages("rvest")
install.packages("quantmod")
install.packages("corpcor")
```

First we get the names of stocks within the ASX200 by scraping the table found on the Wikipedia page <https://en.wikipedia.org/wiki/S%26P/ASX_200>. We use the `rvest` package to use fancy syntax.

```{r}
library("rvest")
url = "https://en.wikipedia.org/wiki/S%26P/ASX_200"
asx200 = url %>%
  read_html() %>%
  html_nodes(xpath='//table') 
asx200[[1]] %>% html_table() -> asx200
```

```{r}
asx200
```

We will download the returns using the `quantmod` package.

```{r include=FALSE}
library("quantmod")
```

We now get a list of the tickers that compose the ASX200.

```{r}
all.tickers = lapply(asx200[,1], function(t) paste(t, ".AX", sep=""))$Code
```

We can print them out.

```{r}
all.tickers
```

We are going to restrict the data downloaded to 5 randomly chosen stocks.

```{r}
tickers = sample(all.tickers, 5)
print(tickers)
```

Alternatively, some crypto-currencies.

```{r}
tickers = as.character(c("BTC-USD", "ETH-USD", "SOL1-USD", "DOGE-USD", "XRP-USD", "APT-USD"))
```

Alternatively, evaluate this next line for a selection of stocks.

```{r}
tickers = as.character(c("BHP.AX", "RIO.AX", "WBC.AX", "CBA.AX", "NAB.AX", "WDS.AX", "SUN.AX", "FMG.AX"))
```

We get prices for all tickers.

```{r}
getSymbols(tickers, from="2022-01-01", to="2022-10-01", src="yahoo", auto.assign = TRUE)
```

We download daily returns and merge them into a single data.frame.

```{r}
daily.returns = do.call(merge, lapply(tickers, function(x) dailyReturn(get(x))))
names(daily.returns) = tickers
```

We now solve for the optimal portfolio.

We will use the sample covariance matrix.

```{r}
S = cov(daily.returns)
round(S, 5)
```

We check correlations.

```{r}
round(cov2cor(S), 2)
```

We will calculate the "plug-in optimal portfolio".

```{r}
mu.hat = colMeans(daily.returns)
mp = markowitz(mu.hat, S)
```

```{r}
mp$R
```

We get the optimal returns (in percent p.a.)

```{r}
mp$R * 100 * 365 # % p.a.
```

We get the optimal portfolio.

```{r}
round(mp$Pi, 4)
```

Construct an equally weighted portfolio.

```{r}
eq.Pi = (mp$Pi*0+1)/length(mp$Pi)
round(eq.Pi, 4)
```

```{r}
t(eq.Pi) %*% mu.hat * 100
```

We plot the value of the portfolio assuming that we have normalised it so the value starts at \$1 on the first day.

```{r}
V.p = cumprod(1 + daily.returns %*% mp$Pi)
plot(time(daily.returns), V.p, type='l', main="Portfolio value in-sample")
```

We get prices again for an out-of-sample test.

```{r}
getSymbols(tickers, from="2022-10-01", src="yahoo")
```

We download daily returns.

```{r}
fwd.daily.returns = do.call(merge, lapply(tickers, function(x){ dailyReturn(get(x)) }))
names(fwd.daily.returns) = tickers
```

We plot the value of the portfolio assuming that we have normalised it so the value starts at \$1 on the first day.

```{r}
fwd.V.p = cumprod(1+fwd.daily.returns %*% mp$Pi)
plot(time(fwd.daily.returns), fwd.V.p, type='l', main="Portfolio value out-of-sample")
```

## With short-selling constraints

If you want to get fancy, we can also solve a similar optimization with a short selling constraint. In other words, we create a "long-only" portfolio. Mathematically, this means that we restrict non-negative weights for the portfolio.

```{r}
N = ncol(S)
aMat = cbind(t(array(1, dim = c(1,N))), diag(N))
b0 = as.matrix(c(1, rep.int(0,N)))
zeros = array(0, dim = c(N,1))

res = solve.QP(S, zeros, aMat, bvec=b0, meq = 1)
```

We get the optimal returns (in percent p.a.)

```{r}
Pi.p.longonly = res$solution
R = Pi.p.longonly %*% mu.hat
R * 100 * 250 # % p.a.
```

We compare the long/short optimal portfolio to the long-only portfolio in-sample.

```{r}
Vl.p = cumprod(1+daily.returns %*% Pi.p.longonly)
plot(time(daily.returns), V.p, type='l', main="Portfolio value in-sample")
lines(time(daily.returns), Vl.p, type='l', lty=2, col='red')
legend("topleft", c("long/short", "long only"), lty=c(1,2), col=c(1,"red"), text.width=40, bty = "n")
```

We compare the long/short optimal portfolio to the long-only portfolio out-of-sample.

```{r}
fwd.Vl.p = cumprod(1+fwd.daily.returns %*% Pi.p.longonly)
plot(time(fwd.daily.returns), fwd.V.p, type='l', main="Portfolio value out-of-sample")
lines(time(fwd.daily.returns), fwd.Vl.p, type='l', lty=2, col='red')
legend("topleft", c("long/short", "long only"), lty=c(1,2), col=c(1,"red"), text.width=40, bty = "n")
```

```{r}
lambdas = eigen(S, only.values = TRUE)$values
plot(lambdas, pch=19, type="b")
```

Mean price change over a year.

```{r}
mu.hat * 100 * 250
```

Covariation in price over a year.

```{r}
round(S * 100 * 250, 2)
```

# Over-prediction of returns: a simulation study

We illustrate the over-prediction phenomena by performing a Monte-Carlo simulation.

## Realistic parameters

We choose some "realistic" parameters for our stock price simulation.

```{r}
avg.yearly.return = 6
avg.yearly.retsd = 4.5
```

## A run-through of the simulation steps

First, we run through the steps that we are going to take in our simulation to make sure that everything works well.

First, we set the number of stocks.

```{r}
p = 10
```

Generate stock names.

```{r}
tickers = as.character(lapply(seq(1,p), function(x) paste("S", x, sep="")))
```

Generate some realistic (population) returns and print the corresponding percentage return per annum.

```{r}
mu = rnorm(p, mean=avg.yearly.return / (100 * 250), sd=avg.yearly.retsd / (100 * 250))
names(mu) = tickers
mu * 100 * 250 # % p.a.
```

We generate a reasonable (population) covariance matrix for daily returns.

```{r}
pcor = function(rho, p) {
  Tn = matrix(0, p, p)
  for (i in 1:p) {
    for (j in 1:p) {
      Tn[i,j] = rho^abs(i-j)
    }
  }
  return(Tn)
}

Sigma = (rWishart(1, df=500, pcor(0.7, p))/(500-1))[,,1]
Sigma = avg.yearly.retsd*Sigma/max(Sigma)
Sigma = Sigma / (100 * 250)
rownames(Sigma) = tickers
colnames(Sigma) = tickers

print(Sigma * 100 * 250, digits=2) # % p.a.
```

We look at the correlations.

```{r}
print(cov2cor(Sigma), digits=2)
```

We calculate the optimal theoretical portfolio based on the (population) $\mu$ and $\Sigma$.

```{r}
mp = markowitz(mu, Sigma)
```

We get the optimal portfolio.

```{r}
Pi = mp$Pi
Pi
```

And the optimal returns.

```{r}
R = mp$R
R * 100 * 250 # % p.a.
```

We now simulate large number of returns for the stocks. We use the `mvtnorm` package so we can take into account $\Sigma$.

```{r}
library(mvtnorm)
```

```{r}
n = 500
Returns = rmvnorm(n, mean=mu, sigma=Sigma)
colnames(Returns) = tickers
```

We plot the simulated stock returns.

```{r}
plot(ts(Returns), type='l', main="Stock Returns")
```

We plot the simulated stock prices, assuming that all stocks start at \$1 price at the initial time. This gives us "realistic" prices over $n=`r n`$ days.

```{r}
Prices = apply(Returns, 2, function(x) cumprod(1+x))
plot(ts(Prices))
```

We now calculate the sample mean.

```{r}
sample.mu = colMeans(Returns)
names(sample.mu) = tickers
sample.mu * 100 * 250 # % p.a.
```

And sample covariance matrix.

```{r}
S = cov(Returns)
rownames(S) = tickers
colnames(S) = tickers
```

We print the correlations.

```{r}
print(cov2cor(S), digits=2)
```

We get the ratio of $p$ to $n$.

```{r}
y = p/n
y
```

We calculate the "plug-in portfolio" based on the sample mean $\bar{\mathbb{x}}$ and sample covariance $\mathbb{S}$.

```{r}
mp = markowitz(sample.mu, S)
```

We get the "plug-in portfolio".

```{r}
Pi.p = mp$Pi
Pi.p
```

And the "plug-in returns".

```{r}
R.p = mp$R
R.p * 100 * 250 # % p.a.
```

## Full simulation

We are going to perform this simulation over the following values of $p$.

```{r}
ps = c(10, 50, 100, 120, 160, 200, 220, 240)
```

For each value of $p$, we are going to perform $M$ simulations and take the average of the plug-in returns.

```{r}
M = 30
```

The number of days for returns.

```{r}
n = 250
```

We now do the full simulation based on the steps above.

```{r}
R = rep(0, length(ps))
avg.Rp = rep(0, length(ps))

for (k in 1:length(ps)) {
  p = ps[k]
  cat('p:', p, '\n')

  # population values for mu and Sigma
  mu = rnorm(p, mean=avg.yearly.return / (100 * 250), sd=avg.yearly.retsd / (100 * 250))
  Sigma = (rWishart(1, df=500, pcor(0.7, p))/(500-1))[,,1]
  Sigma = avg.yearly.retsd*Sigma/max(Sigma)
  Sigma = Sigma / (100 * 250)

  # theoretical return
  mp = markowitz(mu, Sigma)
  R[k] = mp$R
  
  Rp = rep(0, M)
  for (m in 1:M) {
    # simulated returns
    Returns = rmvnorm(n, mean=mu, sigma=Sigma)
    sample.mu = colMeans(Returns)
    S = cov(Returns)
    mp = markowitz(sample.mu, S)
    Rp[m] = mp$R
  }
  
  avg.Rp[k] = mean(Rp)
}
```

```{r}
hist(Rp * 100 * 250, breaks='FD')
```

Plot in terms of % p.a.

```{r}
plot(ps, avg.Rp * 100 * 250, type='l', lty=3)
lines(ps, R * 100 * 250, col='red')
```

# A parametric model for the sample covariance

We will need the Marchenko-Pastur density.

```{r}
dmp = function(x, y, sigmasq=1) {
  a = sigmasq*(1-sqrt(y))^2
  b = sigmasq*(1+sqrt(y))^2
  ifelse(x <= a | x >= b, 0, suppressWarnings(sqrt((x - a) * (b - x))/(2 * pi * sigmasq * x * y)))
}
```

Sample some eigenvalues from a sample covariance matrix.

```{r}
p = 500
n = 1000
Sigma = diag(1, p, p)

eigenvalues = function(x) eigen(x, only.values=T)$values

S = (rWishart(1, (n-1), Sigma)/(n-1))[,,1]
Ls = eigenvalues(S)
```

Plot the histogram of the eigenvalues compared to the MP density.

```{r}
hist(Ls, breaks=30, plot=F) -> h
plot(h$breaks, c(h$density, 0), type="s",  xlab="", ylab="", xlim=c(0, 1.2*max(Ls)), ylim=c(0,1))
curve(dmp(x, y=p/n), 0, 1.2*max(Ls), lty=3, lwd=2, col="darkmagenta", add=TRUE)
```

Plot the Gaussian kernel estimate with bandwidth $h=0.05$ of the eigenvalues compared to the MP density.

```{r}
par(xaxs="i", yaxs="i", cex=0.8, cex.axis=1.0) -> opar
curve(dmp(x, y=p/n, sigmasq=1), 0, 1.2*max(Ls), lty=3, lwd=1, ylim=c(0,2))
d = density(Ls, kernel="gaussian", bw=0.05)
lines(d, col=1, lwd=1)
par(opar)
```

## Inverse cubic density

The function $h$ from Section 5.

```{r}
h_ = function(t, alpha=0.5) {
  c = 2*(1-alpha)^2
  a = 2*alpha - 1
  if (t >= alpha) {
    return(c/((t-a)^3))
  } else {
    return(0)
  }
}
h = Vectorize(h_)
```

```{r}
plot(h, 0, 4)
x = seq(0, 4, length.out=100)
for (alpha in c(0.6, 0.7, 0.8, 0.9, 0.99)) {
  lines(x, h(x, alpha))
}
```

## Generating a population covariance matrix

Create a (population) covariance matrix with different variances down the diagonal that we will try to estimate.

```{r}
p = 500
n = 1000
y = p/n

Sigma = diag(1, p, p)
for (i in 1:5) {
  diag(Sigma)[sample(1:p, 100)] = i
}

print(diag(Sigma))
```

```{r}
eigenvalues(Sigma)
```

```{r}
rotate = function(A) {t(A)[ncol(A):1,]}
image(rotate(Sigma), axes=FALSE)
```

Generate an example sample covariance matrix and calculate its eigenvalues.

```{r}
eigenvalues = function(x) eigen(x, only.values=T)$values
Ls = apply(rWishart(1, (n-1), Sigma)/(n-1), 3, eigenvalues)
```

Plot the histogram of the eigenvalues.

```{r}
hist(Ls, breaks=50, plot=F) -> h
plot(h$breaks, c(h$density, 0), type="s",  xlab="", ylab="", xlim=c(0, 1.2*max(Ls)), ylim=c(0,1))
curve(dmp(x, y=p/n), 0, 1.2*max(Ls), lty=3, lwd=2, col="darkmagenta", add=TRUE)
```

## Fitting a first model

We assume that the eigenvalues can only be located at the natural numbers $x=1,2,\ldots, k$.

```{r}
k = 5 # number of weights
t = seq(1, k) # location of weights
print(t)
```

Using $m$ points outside of the support of the LSD.

```{r}
m = 20 # should be >= p ?
u = seq(-10, 0, length.out = m)
```

The approximate Stieltjes transform based on the assumptions that the eigenvalues are located at the integers.

```{r}
sn = Vectorize(function(u) {
  y = p/n
  -(1-y)/u + 1/n * sum(1/(Ls-u))
})
```

We solve the least-squares problem by creating an objective function to minimise.

```{r}
obj = function(w) {
  y = p/n
  diff = numeric(m)
  k = length(w)
  
  for (j in 1:m) {
    s = sn(u[j])
    
    sumtotal = 0
    for (i in 1:k) {
      sumtotal = sumtotal + w[i]*t[i]/(1+s*t[i])
    }
    
    diff[j] = 1/s+u[j] - y*sumtotal
  }
  
  sum(abs(diff)^2)
}
```

Solve the constrained problem and extract our estimated weights $\hat w$.

```{r}
theta = c(0.5,0.5,0.5,0.5,0.5) # initial guess

ui = rbind(diag(k), rep(1,k))
ci = c(rep(0,k), 1)

sol = constrOptim(theta, obj, NULL, ui=ui, ci=ci)

wh = sol$par

options(scipen=999)
print(wh, digits=2)
```

A linear combination of discrete weights gives a linear combination of MP densities with different $\sigma^2$.

```{r}
phi = function(x, w) {
  density = 0*x # handle vector 'x' case
  for (i in 1:length(w)) {
    density = density + w[i] * dmp(x, y=p/n, sigmasq=i)
  }
  return(density)
}
```

Plot the density against the histogram.

```{r}
par(xaxs="i", yaxs="i", cex=0.8, cex.axis=1.0) -> opar
hist(Ls, breaks=50, plot=F) -> h
plot(h$breaks, c(h$density, 0), type="s",  xlab="", ylab="", xlim=c(0, 1.2*max(Ls)), ylim=c(0,1))
curve(phi(x, wh), from=0, to=1.2*max(Ls), lty=3, lwd=2, col="darkgreen", add=TRUE)
curve(dmp(x, y=p/n), 0, 1.2*max(Ls), lty=1, lwd=1, col="darkmagenta", add=TRUE)
par(opar)
```

## Second model: solving for weight and location

We are going to solve for the position of the eigenvalues as well as the weights. This means that we are solving for $(2k-1)$ parameters $$
\theta = (\lambda_1, \ldots, \lambda_k, w_1, \ldots, w_{k-1})
$$ and $w_k = 1-\sum_{i=1}^{k-1} w_i$. This means our initial guess will look something like:

```{r}
theta = c(0.5, 1, 2, 3, 4, 5, 6, # positions
          rep(0.1, 6))           # weights
```

We solve the least-squares problem by creating an objective function to minimise.

```{r}
obj = function(theta) {
  y = p/n
  
  k = as.integer( (length(theta)+1)/ 2 )
  t = theta[1:k] # extract locations / positions
  w = theta[(k+1):length(theta)] # extract weights
  w = c(w, 1-sum(w))

  m = 30
  u = c(seq(min(t)-10, min(t)-1, length.out = m/2),
        seq(max(t)+1, max(t)+10, length.out = m/2))
  
  diff = numeric(m)
  for (j in 1:m) {
    s = sn(u[j])
    sumtotal = 0
    for (i in 1:k) {
      sumtotal = sumtotal + w[i]*t[i]/(1+s*t[i])
    }
    diff[j] = 1/s+u[j] - y*sumtotal
  }
  
  sum(abs(diff)^2)
}
```

```{r}
obj(theta)
```

```{r}
k = length(theta)
q = as.integer((k+1)/2)

ui = rbind(diag(k), 
           c(rep(0, q), rep(1, q-1)),
           c(rep(0, q), rep(-1, q-1))
           )
ci = c(rep(0,k), 0, -1)
```

```{r}
ui %*% theta - ci
```

```{r}
sol = constrOptim(theta, obj, NULL, ui=ui, ci=ci)
theta.hat = sol$par
```

```{r}
lambda.hat = theta.hat[1:q] # extract locations
w.hat = theta.hat[(q+1):length(theta)] # extract weights
print(sum(w.hat))
w.hat = c(w.hat, 1-sum(w.hat))

options(scipen=999)
print(lambda.hat, digits=2)
print(w.hat, digits=2)
```

A linear combination of discrete weights gives a linear combination of MP densities with different $\sigma^2$.

```{r}
phi = function(x, lambda, w) {
  density = 0. * x # handle vector 'x' case
  for (i in 1:length(lambda)) {
    density = density + dmp(x, y=p/n, sigmasq=lambda[i])
  }
  return(density/length(lambda))
}
```

```{r}
x = seq(0, 12, length.out=100)
plot(x, phi(x, lambda.hat, w.hat), type='l')
```

Plot the density against the histogram.

```{r}
hist(Ls, breaks=100, plot=F) -> h
plot(h$breaks, c(h$density, 0), type="s",  xlab="", ylab="", xlim=c(0, 1.2*max(Ls)), ylim=c(0,1))
curve(phi(x, lambda.hat, w.hat), from=0, to=1.2*max(Ls), lty=3, lwd=2, col="darkgreen", add=TRUE)
```

## Final attempt: ensuring ordering of eigenvalues

```{r}
q = 10
ub = 10

theta = c(seq(0.1, ub, length.out=q),  # positions
          rep(0.1, q-1))   # weights

theta
```

```{r}
k = length(theta)
q = as.integer((k+1)/2)
```

```{r}
A = matrix(0, q-1, k)
A[1,1] = 1
for (i in 2:(q-1)) {
  A[i,i+1] = 1
  A[i,i] = -1
}
A %*% theta
```

```{r}
ui = rbind(diag(k), 
           c(rep(0, q), rep(1, q-1)),
           c(rep(0, q), rep(-1, q-1)),
           A
           )
ci = c(rep(0,k), 0, -1, rep(0, q-1))
```

```{r}
ui %*% theta - ci
```

```{r}
sol = constrOptim(theta, obj, NULL, ui=ui, ci=ci)
theta.hat = sol$par
```

```{r}
lambda.hat = theta.hat[1:q] # extract locations
w.hat = theta.hat[(q+1):length(theta)] # extract weights
w.hat = c(w.hat, 1-sum(w.hat))

options(scipen=999)
print(lambda.hat, digits=2)
print(w.hat, digits=2)
```

Plot the density against the histogram.

```{r}
hist(Ls, breaks=100, plot=F) -> h
plot(h$breaks, c(h$density, 0), type="s",  xlab="", ylab="", xlim=c(0, 1.2*max(Ls)), ylim=c(0,1))
curve(phi(x, lambda.hat, w.hat), from=0, to=1.2*max(Ls), lty=3, lwd=2, col="darkgreen", add=TRUE)
```
