---
title: "Workshop - Week 8"
output:
  word_document:
    toc: yes
  html_document:
    fig_height: 4
    toc: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{=tex}
\newcommand{\X}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\S}{\mathbf{S}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\C}{\mathbf{C}}
```
# Correlation coefficient

## Population correlation in bivariate case

```{r}
Sigma = matrix(c(5,4,4,5), ncol=2)
Sigma
```

Use `cov2cor` to scale the covariance matrix $\Sigma$ by its diagonal to
become the correlation matrix $$
  R = \begin{pmatrix}1 & \rho\\ \rho & 1 \end{pmatrix}
$$

```{r}
R = cov2cor(Sigma)
R
```

Alternatively, we can compute $\rho$ explicitly in the bivariate case
from the formula
$$ \rho = \frac{\sigma_{12}}{\sqrt{\sigma_{11} \sigma_{22}}}. $$

```{r}
rho = Sigma[1,2]/(sqrt(Sigma[1,1]*Sigma[2,2]))
rho
```

## Sample correlation distribution in bivariate case

We load the `mvtnorm` library for sampling from the multivariate normal
distribution.

```{r}
library(mvtnorm)
```

We sample $N$ observations from a $\mathcal{N}(\mu, \Sigma)$
distribution.

```{r}
N = 100
mu = c(0,0)
Sigma = matrix(c(5,4,4,5), ncol=2)
X = rmvnorm(N, mean=mu, sigma=Sigma)
```

```{r}
dim(X)
```

We can calculate the sample covariance matrix "by hand".

```{r}
S = t(X) %*% X / (N-1)
S
```

Or use the in-built function (which may be a bit more accurate).

```{r}
S = cov(X)
S
```

Then the sample correlation matrix can be obtained by converting the
sample covariance.

```{r}
cov2cor(S)
```

Or alternatively, by calculating it directly from the observations.

```{r}
cor(X)
```

The sample correlation between the coordinates $X_1$ and $X_2$ is $$
R = \frac{S_{12}}{\sqrt{S_{11} S_{22}}}
$$

```{r}
R = S[1,2]/sqrt(S[1,1]*S[2,2])
R
```

*Checking Theorem on Page 4 of lecture*

We now look at the distribution in the case that $\rho = 0$ by sample
$N$ observations from a $N_2(\mu, I)$ distribution.

```{r}
p = 2
Sigma = diag(1, p, p)
mu = rep(0, p)

M = 5000 # number of MC simulations
n = 1000 # number of observations

MC = numeric(M)
for (i in 1:M) {
  X = rmvnorm(n, mean=mu, sigma=Sigma)
  R = cor(X)[1,2]
  MC[i] = R*sqrt((n-1)/(1-R**2))
}
```

Generate histogram of MC simulations.

```{r}
hist(MC, breaks="FD", plot=FALSE) -> h
```

Create a custom histogram plot that looks nice.

```{r}
par(xaxs="i", yaxs="i", cex=0.8, cex.axis=1.0) -> opar

# theoretical density
f = function(x) dt(x, df=n-1)

plot(h$breaks, c(h$density, 0), type="s", xlab="", ylab="", ylim=c(0, 0.5))

curve(f, -3, 3, lwd=1, lty=3, add=TRUE)

title(main="Histogram compared to Student's t density", outer=T, line=-2)

par(opar)
```

## Classic asymptotic

```{r}
p = 2
rho = 0.1 # population correlation is known
Sigma = diag(1, p, p)
Sigma[1,2] = rho
Sigma[2,1] = rho
Sigma
```

```{r}
mu = rep(0, p)

M = 5000 # number of MC simulations
n = 500 # number of observations

MC = numeric(M)
for (i in 1:M) {
  X = rmvnorm(n, mean=mu, sigma=Sigma)
  R = cor(X)[1,2]
  MC[i] = sqrt(n-1)*(R-rho)/(1-rho**2)
}
```

```{r}
hist(MC, "FD", freq=FALSE)
curve(dnorm, col="red", lwd=3, add=TRUE)
```

# Multiple correlation coefficient

## Population multiple correlation

We look at the population multiple correlation coefficient between $X_1$
and $\mathbb{X}_2 = (X_2, X_3, \ldots, X_p)'$. Let's look at a simple
example where $p=5$. Generate a random covariance matrix.

```{r}
p = 5
X = matrix(runif(p*p), ncol=p)
Sigma =  t(X) %*% X
Sigma
```

Partitioning the covariance matrix $\Sigma$ as $$
\Sigma =
\begin{pmatrix}
\sigma_{11} & \bar{\sigma}_{21}'\\
\bar{\sigma}_{21} & \Sigma_{22}
\end{pmatrix}.
$$ The multiple correlation coefficient between $X_1$ and $\mathbf{X}_2$
is $$
\rho =
\sqrt{\frac{\bar{\sigma}_{21}' \Sigma_{22}^{-1}\bar{\sigma}_{21}}{\sigma_{11}}}
$$

```{r}
t(Sigma[2:p,1])
```

```{r}
dim(Sigma[2:p,2:p])
```

```{r}
Sigma
```

```{r}
Sigma[2:p, 2:p]
```

```{r}
p = ncol(Sigma)
rho = sqrt( t(Sigma[2:p,1]) %*% solve(Sigma[2:p,2:p]) %*% Sigma[2:p,1] / Sigma[1,1] )
rho
```

## Sample multiple correlation

We sample $N$ observations from a $\mathcal{N}(\mu, \Sigma)$
distribution.

```{r}
N = 100
mu = rep(0, p)
X = rmvnorm(N, mean=mu, sigma=Sigma) # data matrix
```

```{r}
dim(X)
```

The sample MCC is given by

```{r}
S = cov(X)
p = nrow(S)
R = sqrt(t(S[2:p,1]) %*% solve(S[2:p,2:p]) %*% S[2:p,1] / S[1,1])
R
```

## Sample multiple correlation distribution

We perform a simulation.

```{r}
p = 5
Sigma = diag(1, p, p)
mu = rep(0, p)

M = 8000 # number of MC simulations
n = 1000 # number of observations

MC = numeric(M)
for (i in 1:M) {
  X = rmvnorm(n, mean=mu, sigma=Sigma) 
  S = cov(X)
  R = sqrt(t(S[2:p,1]) %*% solve(S[2:p,2:p]) %*% S[2:p,1] / S[1,1])
  Rsq = R**2
  MC[i] = (n-p)/(p-1) * Rsq/(1-Rsq)
}
```

Generate histogram of MC simulations.

```{r}
hist(MC, breaks="FD", plot=FALSE) -> h
```

Create a custom histogram plot that looks nice.

```{r}
par(xaxs="i", yaxs="i", cex=0.8, cex.axis=1.0) -> opar

# theoretical density
f = function(x) df(x, df1=p-1, df2=n-p) # compared with F distribution

plot(h$breaks, c(h$density, 0), type="s", xlab="", ylab="", ylim=c(0, 0.9))

curve(f, 0, 5, lwd=1, lty=3, add=TRUE)

title(main="Histogram compared to F density", outer=T, line=-2)

par(opar)
```

## Sample multiple correlation distribution - Adjusted case

We perform a simulation.

```{r}
p = 5
Sigma = diag(1, p, p)
mu = rep(0, p)

M = 5000 # number of MC simulations
n = 1000 # number of observations

MC = numeric(M)
for (i in 1:M) {
  X = rmvnorm(n, mean=mu, sigma=Sigma)
  S = cov(X)
  R = sqrt(t(S[2:p,1]) %*% solve(S[2:p,2:p]) %*% S[2:p,1] / S[1,1])
  Rsq = R**2 - (p-1)/(n-p)*(1-R**2) # Adjusted version!
  MC[i] = (n-p)/(p-1) * Rsq/(1-Rsq)
}
```

Generate histogram of MC simulations.

```{r}
hist(MC, breaks="FD", plot=FALSE) -> h
```

Create a custom histogram plot that looks nice.

```{r}
par(xaxs="i", yaxs="i", cex=0.8, cex.axis=1.0) -> opar

# theoretical density
f = function(x) df(x, df1=p-1, df2=n-p)

plot(h$breaks, c(h$density, 0), type="s", xlab="", ylab="", ylim=c(0, 0.9))

curve(f, 0, 5, lwd=3, lty=3, add=TRUE)

title(main="Histogram compared to F density", outer=T, line=-2)

par(opar)
```

Notice that the theorem in the lecture doesn't apply to the adjusted
R-squared given by $R^*$.

# High-dimensional regime / MP regime

## Over-estimation of $R^2$

Choose a large $p$.

```{r}
p = 500
```

Generate a population covariance matrix, AR1-style covariance.

```{r}
pcor = function(rho, p) {
  Tn = matrix(0, p, p)
  for (i in 1:p) {
    for (j in 1:p) {
      Tn[i,j] = rho^abs(i-j)
    }
  }
  return(Tn)
}

Sigma = pcor(0.6, p)
```

Calculate the population MCC.

```{r}
p = ncol(Sigma)
rho = sqrt(t(Sigma[2:p,1]) %*% solve(Sigma[2:p,2:p]) %*% Sigma[2:p,1] / Sigma[1,1])
rho
```

```{r}
mu = rep(0, p)

M = 500 # number of MC simulations
n = 2*p # number of observations
y = p/n

MC = numeric(M)
for (i in 1:M) {
  X = rmvnorm(n, mean=mu, sigma=Sigma)
  S = cov(X)
  R = sqrt(t(S[2:p,1]) %*% solve(S[2:p,2:p]) %*% S[2:p,1] / S[1,1])
  Rsq = R**2
  MC[i] = Rsq - (1-y)*rho**2-y
}
```

We can see that the quantity $R^2 - (1-y)\rho^2 - y$ is closely
distributed around zero, or in other words, $R^2$ does not converge to
$\rho^2$ as desired! So in the high-dimensional case,
$R^2 \to (1-y)\rho^2 + y$ instead compared to the classic theory that
states $R^2 \to \rho^2$.

```{r}
mean(MC)
```

```{r}
sd(MC)
```

## Central limit theorem

Choose a large $p$.

```{r}
p = 250
```

Generate a population covariance matrix.

```{r}
pcor = function(rho, p) {
  Tn = matrix(0, p, p)
  for (i in 1:p) {
    for (j in 1:p) {
      Tn[i,j] = rho^abs(i-j)
    }
  }
  return(Tn)
}

Sigma = pcor(0.6, p)
```

Calculate the population MCC.

```{r}
p = ncol(Sigma)
rho = sqrt(t(Sigma[2:p,1]) %*% solve(Sigma[2:p,2:p]) %*% Sigma[2:p,1] / Sigma[1,1])
rho
```

```{r}
mu = rep(0, p)

M = 1000 # number of MC simulations
n = 500 # number of observations
y = p/n

MC = numeric(M)
for (i in 1:M) {
  X = rmvnorm(n+1, mean=mu, sigma=Sigma)
  S = cov(X)
  R = sqrt(t(S[2:p,1]) %*% solve(S[2:p,2:p]) %*% S[2:p,1] / S[1,1])
  MC[i] = sqrt(n)*(R^2-y-(1-y)*rho^2)
}
```

Generate histogram of MC simulations.

```{r}
hist(MC, breaks="FD", plot=FALSE) -> h
```

```{r}
sigmasq = function(t) 2*(y+(1-y)*t)^2 - 2*(-2*(1-y)*t^2 + 4*(1-y)*t+2*y)*(y+(1-y)*t-0.5)
```

Create a custom histogram plot that looks nice.

```{r}
par(xaxs="i", yaxs="i", cex=0.8, cex.axis=1.0) -> opar

# theoretical density
sd = sqrt(sigmasq(rho^2))
f = function(x) dnorm(x, mean=0, sd=sd)

plot(h$breaks, c(h$density, 0), type="s", xlab="", ylab="", ylim=c(0, 1.0))
curve(f, -5, 5, lwd=1, lty=3, add=TRUE)

title(main="Histogram compared to theoretical density", outer=T, line=-2)

par(opar)
```
